---
title: "Gross Domestic Product Analysis"
author: "Jean Jecha"
date: "June 20, 2016"
output:
  html_document:
    keep_md: true
---
### GDP Data Analysis

This report is an analysis of the gross domestic product (GDP) ranking data with education statistics for countries. Some of the analysis includes GDP summaries by the two high income groups. Graph showing totals by income groups. Breaking up the GDP rankings into separate quantiles and analyzing that with the education income groups. Finally viewing the top 38 ranking nations that are lower middle income. 


```{r Setup data}
########################################
#  Load packages needed for analysis
########################################
# Load packages
library(repmis)
library(countrycode)
library(plyr)
library(dplyr)
library(ggplot2)

####################################################
#   Dowload and clean data files
####################################################
#  Place the GDP data URL into the object FinURL
FinURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"

#  Download data and read into data frame
FGDPData <- read.csv(FinURL,skip = 5, header=FALSE,sep = ',' )
# str(FGDPData)
# head(FGDPData)

#  Get Usable columns and remove rows with no country code or rank
#    get columns 1 - 6 and remove rows with no ranking data
FGDPData.2 <- FGDPData[!FGDPData$V2 == "",c(1:6)]

#  Final columns of country code, ranking, country and GDP amount
#      Eliminate any rows with no country code
FGDPData.3 <- FGDPData.2[!FGDPData.2$V1 == "",c(1:2,4:5)]

#  adding column headers
col_headings <- c("CountryCode", "Ranking", "country", "GDP")
names(FGDPData.3) <- col_headings

#  change rankink column to integer for sorting
FGDPData.3$Ranking <- as.integer(as.character(FGDPData.3$Ranking))

#  remove commas and convert to numeric, 
#    needed to remvove commas for conversion to numeric
FGDPData.3$GDPAmount <- gsub(",","",FGDPData.3$GDP)
FGDPData.3$GDPAmount <- as.numeric(as.character(FGDPData.3$GDPAmount))

#  convert country code to character
FGDPData.3$CountryCode <- as.character(FGDPData.3$CountryCode)


#  Place the education URL into the object FedStatsURL
FedStatsURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"

#  Download data and read into data frame
FedStatsData <- read.csv(FedStatsURL, header=TRUE,sep = ',' )

#  converting income group and country code to character
FedStatsData$IncomeGroup <- as.character(FedStatsData$Income.Group)
FedStatsData$CountryCode <- as.character(FedStatsData$CountryCode)
#  str(FedStatsData)

#  removing country codes with no income groups
FedStatsData.2 <- FedStatsData[!FedStatsData$Income.Group == "",c(1,32)]

```


#### Countries in GDP matching countries in educatianal data
The following shows the number of rows that match in both the GDP and educational data for both files. This is after the data is cleaned and matched.
```{r MatchingCountryCode}
# Question 1 row count of matching rows
length(match(FGDPData.3$CountryCode, FedStatsData.2$CountryCode, nomatch = 0))

# merge GDP with educational data
MergeGDPData <- merge(
  FGDPData.3, FedStatsData.2,
  by="CountryCode",
  all = TRUE
)

#  Remove NA rows due to merging files
GDPDataTidy <- MergeGDPData[is.na(MergeGDPData$GDPAmount)==0,]
GDPDataTidy <- na.omit(MergeGDPData)

```


##### Reporting the 13th country from the lower ranked 
Based on the GDP Ranking from lowest ranking to highest ranking the following country is 13th in the rank. With the United States ranked highest their rank would be last in the list before retrieving the 13th country. 
```{r country13}
# Question 2, sort GDP rank listing the 13th country 
Row13 <- GDPDataTidy[order(GDPDataTidy$Ranking,decreasing = TRUE),]
Row13[13,]

```


##### Average GDP Amount
Analysis of the average GDP Amount in the income groups of high income: OECD and high income: nonOECD
```{r AvgGDP}
# Question 3 Average GDP rankings for High income OECD and High Income nonOECD
GDPHighIncome <- GDPDataTidy[
      GDPDataTidy$IncomeGroup == "High income: nonOECD" | 
      GDPDataTidy$IncomeGroup == "High income: OECD", ]

GDPHighIncome.2 <- GDPHighIncome[,5:6]
aggregate(GDPHighIncome.2[, 1], list(GDPHighIncome.2$IncomeGroup), mean)
```


##### GDP graph by Income Group
Graph indicating the GDP income broken out by the country income group.
```{r IncomeGroupPlot}
#  Question 4 ggplot2
ggplot(data=GDPDataTidy, aes(x=IncomeGroup, fill = IncomeGroup)) + 
    geom_bar()
```


##### GDP ranking quantile compared to Income Group
Breaking out GDP ranking into quantile groups to compare with the country income group. Table showing the quantile groups with income group and a count of countries within these pairs.
```{r QuantileRank}
#  Question 5 part 1 
#   Breaking into quantiles showing results vs. Income Group
GDPDataTidyQuantile <- GDPDataTidy %>% mutate(quantile = ntile(Ranking, 5))
ddply(GDPDataTidyQuantile, c("quantile", "IncomeGroup"),
      summarise,NbrOccur=length(GDPAmount))
```


##### Top 38 ranked countries with lower middle income
Results of the top 38 ranked countries that fall into the lower middle income range of education.
```{r LowerMiddle38}
#  Question 5 part 2 
#    Showing top 38 GDP ranked countries with "Lower Middle Income"
GDP38 <- GDPDataTidy[order(GDPDataTidy$Ranking,decreasing = FALSE),]
GDP38 <- GDP38[1:38,]
GDP38[GDP38$IncomeGroup == "Lower middle income",c(1:3,5:6)]
```

##### Summary
To conclude this analysis the final data contained 190 GDP countries. The High Income: OECD group has a GDP average of 1,483,917 compared to the nonOECD average of 104,349. The Lower Middle Income group has the largest GDP total and Upper Middle Income being second largest GDP. Five of the top 38 GRP ranking countries are in the Lower Middle income group. 